---
title: "01-02-functions-intro"
output: html_document
editor_options:
  chunk_output_type: console
---

### Why Functions?

Functions allow us to:
- avoid repeating code
- make programs easier to read
- break complex problems into smaller pieces
- reuse logic across projects

At a high level, a function:
- takes **inputs**
- performs **operations**
- returns **output**

---

### A Very Simple Function

Let’s start with the simplest possible function.

```{r}
add_two_numbers <- function(x, y) {
  x + y
}
```

```{r}
add_two_numbers(2, 3)
```

This function:
- takes two inputs: `x` and `y`
- returns their sum

---

### Function Inputs and Outputs

Functions can take any number of inputs.

```{r}
square_number <- function(x) {
  x^2
}
```

```{r}
square_number(4)
```

#### Student practice

```{r exercise-square}
#| eval: false
# Write a function that cubes a number
cube_number <- function(x) {
  ___
}

cube_number(___)
```

---

### Default Arguments

Functions can also have **default values**.

```{r}
power <- function(x, p = 2) {
  x^p
}
```

```{r}
power(3)
power(3, 3)
```

This allows flexibility while keeping sensible defaults.

#### Student practice

```{r exercise-default_arg}
#| eval: false
# Modify the function so the default power is 3
power <- function(x, p = ___) {
  x^p
}

power(2)
power(2, 4)
```

---

### Returning Values Explicitly

R automatically returns the last expression, but we can also use `return()`.

```{r}
absolute_difference <- function(x, y) {
  diff <- x - y
  return(abs(diff))
}
```

```{r}
absolute_difference(10, 3)
```

---

### Functions Inside Functions

Functions can call other functions.

```{r}
mean_of_squares <- function(x) {
  squares <- x^2
  mean(squares)
}
```

```{r}
mean_of_squares(c(1, 2, 3, 4))
```

#### Student practice

```{r exercise-nested_function}
#| eval: false
# Write a function that returns the mean of cubed values
mean_of_cubes <- function(x) {
  ___
}

mean_of_cubes(c(1, 2, 3))
```

---

### Writing Cleaner Code with Functions

Functions help keep your analysis clean and readable.

```{r}
calculate_growth_rate <- function(start, end) {
  (end - start) / start
}
```

```{r}
calculate_growth_rate(100, 125)
```

This pattern becomes especially important as projects grow.

---

### Common Mistakes with Functions

A few things to watch out for:
- forgetting to return a value
- using variables that exist outside the function
- mismatching argument names

```{r}
bad_function <- function(x) {
  y <- x + 2
}
```

```{r}
bad_function(5)
```

Notice that this function does not return anything useful.

#### Student practice

```{r exercise-fix_function}
#| eval: false
# Fix the function so it returns a value
fixed_function <- function(x) {
  y <- x + 2
  ___
}

fixed_function(5)
```

---

### Key Takeaways

- Functions encapsulate logic
- Inputs go inside parentheses
- Code lives inside `{ }`
- Functions return values
- Writing functions early saves time later

We will build on this idea extensively throughout the course.
