---
title: "01-01-intro"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Heading 1

## Heading 2

### R as a *Sophisticated* Calculator

This is *text* and this is **bold**

```{r exercise-calc}
2+2
```

```{r exercise-calc}
exp(4)
```

```{r exercise-calc}
1 + 2

exp(1)

log(exp(1))

2^3
```

### Vectors

`r` is a vectorised language. Vector is a collection of objects. R can do operations on the element level in a vector. A vector is created using the `c()` where elements are separated by a comma.

------------------------------------------------------------------------

```{r exercise-vectors}
x <- c(1, 3, 4, 8, 16, 31.99 , 5 )

sum(x)
x
str(x)
length(x)

 x1 <- 5*x
 
 x1
 
 5*x +2
 
 # These are comments
 # we can generate a sequence and then do computations
 
 x_1 <- 1:7 * x* 7:1
 
 log(x)
```

### Some other useful code snippets

```{r}
###################################################
### subsetting
###################################################
x[c(1, 4)]
x[-c(2, 3, 5)]

x[c(2, 3, 5)]

###################################################
###  patterned
###################################################
ones <- rep(1, 10)
ones
rep1 <- print(rep(c(1,2,3), each =2))

rep2 <- print(rep(c(1,2,3), 2))

rep3 <- print(rep(c(1,2,3), c(3,2,1)))
rep3

 seq(from = 2, to = 20, by = 7)


trend <-   1981:2005
trend
rm(trend)
```

### Generating random numners

```{r}
x <- runif(10,1, 100)
?runif()
x <- runif(max = 100, min = 1, n = 10)
x <- runif(100,  1, 10)
set.seed(9876) # this helps reproduce examples
x <- runif(10,1, 100)
x
x <- ceiling(x) #rounds up to the nearest integer. 
x #Note 13.27 becomes 14 (not 13)
```


### R Objects

Let us work with some data and packages. Packages are user written codes with extend R's functionality beyond the base version and make it one of the most powerful tools for data analysis. New packages need to be installed first but once they are installed on your system, they just need to be loaded once per session. All these packages are already loaded in the workspace. If not, you would run the  command  `install.packages("package_name")` for each package

```{r exercise-data_view}


library(tidyverse)
library(dplyr)
library(gapminder)
library(hablar)
library(skimr)

life.expt <- gapminder::gapminder


# or

#View(life.expt)

#first few rows

head(life.expt)

# last few rows
tail(life.expt)

#This is an impportant command to understand your data
str(life.expt)

#some quick summary stats
skim(life.expt)
glimpse(life.expt)

#another useful function about the type of data
class(life.expt)
# tibble vs a dataframe
print(life.expt)

#We can access columns with the $ function

# mean(lifeExp)
mean(life.expt$lifeExp)

#Let us convert regions to factors and gdp as numeric
str(life.expt)

life.expt <- life.expt %>% 
        convert(fct(continent),
                num(year))
str(life.expt)
class(life.expt$continent)
```

### Indexing and subsetting

R has some really useful tools for indexing. Notice the difference between indexing with \[\] vs the accessor function `$`. With the latter, `R` simplifies the dimension, while in the first case, we get back another tibble.

```{r}
#Get the 4th column
str(life.expt[,4])
str(life.expt$lifeExp)

sort(life.expt[,4]) # does not work

#sort is a function that works on vectors and not on data frames or tibbles
sort(life.expt$lifeExp)

# Same as life.expt$lifeExp

# we can sort it.
sort(life.expt$lifeExp)

# Does not help us in identifying any country.

```

AS we saw above, `sort()` will simply sort the data and `order()` provides the index from column. But suppose we want to find the life expectancy rank of the first five countries in alphabetical order. Also, `order(-x)` provides order in descending order.

```{r }

# Add a new column
life.expt$ranking <- rank(-life.expt$lifeExp)

# Next, we get ordering applied to country names

#  invisible() > results will not be printed
invisible(sort(life.expt$country))

life.expt[1:5, c(1,4,7)]
```

Let's see only the countries from Africa who have much lower life expectancy (less than 45 years)

```{r }
africa <- life.expt$continent =="Africa"
poor.health <- life.expt$lifeExp < 45

ind <- africa & poor.health # this is a logical function. Returns TRUE or FALSE
ind
life.expt[ind, ]
```

**Get's the job done** but later on we will see how easy it is to perform these operations using the ***tidyverse*** approach!
