---
title: "02-04-mlbasics"
format: html
editor: visual
---

## Basics of setting up the data for ML

### Set up the libraries

```{r}
#| echo: true
#| output: false
#| 
library(tidyverse)
library(readr)

library(caret)
library(mlbadata)
library(fastDummies)
library(broom)
```

------------------------------------------------------------------------

## Import the dataset

``` {r}        
housing_df <- mlbadata::WestRoxbury
str(housing_df)
```

### Inspect variable names

``` {r}          
names(housing_df)
```

``` {r}          
t(t(names(housing_df)))
```

------------------------------------------------------------------------

## Creating factor and dummy variables

The `REMODEL` column is a character variable. We can convert it to a factor variable.

``` {r}          
housing_df$REMODEL <- factor(
  housing_df$REMODEL,
  levels = c("None", "Recent", "Old")
)
```

```{r}           
str(housing_df)
```

### Student exercise: factor variables

```{r}           
#| eval: false
# Convert the column REMODEL to a factor
# Set the levels in the order:
# None, Recent, Old
# Inspect the structure afterward
```

------------------------------------------------------------------------

## Creating dummy variables (one-hot encoding)

Several algorithms require converting factor variables into indicator columns.\
We use `dummy_columns()` from the **fastDummies** package.

``` {r}          
args(dummy_columns)
```

The key arguments are:

-   remove_selected_columns

-   remove_first_dummy

-   remove_most_frequent_dummy

### One-hot encoding (keep all categories)

``` {r}          
housing_df1 <- housing_df |> 
  dummy_columns(remove_selected_columns = TRUE)
```

``` {r}          
str(housing_df1)
```

### Avoiding the dummy variable trap

In regression models, we must drop one category.

``` {r}          
housing_df2 <- housing_df |> 
  dummy_columns(
    remove_selected_columns = TRUE,
    remove_first_dummy = TRUE
  )
```

``` {r}          
str(housing_df2)
```

### Student exercise: dummy variables

``` {r}          
#| eval: false
# Create dummy variables for housing_df
# Remove the original factor columns
# Remove the first dummy to avoid multicollinearity
```

---
## Data partitioning: training and test sets

We will create an 80–20 split using **caret**.

``` {r}          
set.seed(1234)

trainind <- createDataPartition(
  housing_df$TOTAL.VALUE,
  p = 0.8,
  list = FALSE
)
```

``` {r}          
train_df <- housing_df2[trainind, ]
test_df  <- housing_df2[-trainind, ]
```

The training set will be used to build the model and the test set to evaluate performance.

### Student exercise: partitioning

``` {r}          
#| eval: false
# Use createDataPartition() to:
# Split the data into 75% training and 25% testing
# Use TOTAL.VALUE as the outcome variable
```

------------------------------------------------------------------------

## Basic multiple regression

We use the formula interface:

``` {r}          
outcome ~ predictors
```

``` {r}          
reg1 <- lm(
  TOTAL.VALUE ~ LOT.SQFT + LIVING.AREA,
  data = housing_df2
)
```

```{r}           
summary(reg1)
```

``` {r}          
tidy(reg1)
```

Fit statistics can be obtained using `glance()`.

```{r}           
reg1 |> glance()
```

---
## Interaction effects

```  {r}         
reg2 <- lm(
  TOTAL.VALUE ~ LOT.SQFT * LIVING.AREA,
  data = housing_df2
)
```

Let's view the results:

``` {r}          
reg2 |> tidy()
```

``` {r}          
reg2 |> glance()
```

If we want **only** the interaction term:

``` {r}          
lm(
  TOTAL.VALUE ~ LOT.SQFT:LIVING.AREA,
  data = housing_df2
) |> tidy()
```

### Student exercise: interactions

```  {r}         
#| eval: false
# Fit a regression with:
# TOTAL.VALUE as outcome
# LOT.SQFT, LIVING.AREA, and their interaction
# Inspect the coefficients
```

---
## Including and excluding predictors

Using `.` includes all predictors.

``` {r}          
lm(
  TOTAL.VALUE ~ . -YR.BUILT -TAX,
  data = housing_df2
) |> tidy()
```

------------------------------------------------------------------------

## Fitting a model on training data

``` {r}          
reg3 <- lm(
  TOTAL.VALUE ~ . -YR.BUILT -TAX,
  data = train_df
)
```

The model object contains:

-   coefficients

-   fitted values

-   residuals

We can attach these to the dataset using `augment()`.

``` {r}          
train_predict <- augment(reg3)
```

``` {r}          
train_predict |> 
  select(TOTAL.VALUE, .fitted, .resid)
```

---
## Predicting on test data

### Option 1: `predict()`

``` {r}          
test_pred <- predict(reg3, newdata = test_df)
head(test_pred)
```

### Option 2: `augment()`

``` {r}          
test_pred2 <- augment(reg3, newdata = test_df)
str(test_pred2)
```

---
## Student exercises: ML workflow practice

```   {r}        
#| eval: false
# Fit a regression model using train_df
# Predict TOTAL.VALUE on test_df
# Compare predicted vs actual values
```

``` {r}          
#| eval: false
# Using reg3:
# Extract fitted values and residuals
# Create a small dataframe with:
# actual value, fitted value, residual
```

