---
title: "04-01-summ-cor"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning and Exploratory Diagnostics

### Learning objectives

* Convert binary variables into modeling-ready factors
* Generate multiple exploratory plots programmatically
* Compare loop-based and functional approaches to plotting
* Arrange multiple ggplot objects on a single page
* Use visualization to inform modeling decisions

---

## Load required libraries

```{r}
library(tidyverse)
library(caret)
library(corrplot)
library(psych)
library(hablar)
library(skimr)
library(purrr)
library(readr)
library(patchwork)
library(mlbadata)
```

---

## Converting variables to factors

```{r}
housing.df <- mlbadata::BostonHousing

cols <- c("CHAS", "CAT.MEDV")

housing.df[cols] <- map(
  housing.df[cols],
  factor,
  levels = c(1, 0),
  labels = c("Yes", "No")
)

head(housing.df)
```

```{r}
glimpse(housing.df)
```

---

## Generating multiple exploratory plots

```{r}
selected.vars <- c("MEDV", "CHAS", "RM", "ZN", "PTRATIO", "AGE", "CRIM")
df1 <- housing.df[, selected.vars]
```

---

## Loop-based density plots (solved)

```{r}
for (i in 3:ncol(df1)) {
  print(
    df1 |> 
      ggplot(aes(x = df1[, i], color = CHAS)) +
      geom_density() +
      labs(x = colnames(df1[i]))
  )
}
```

---

### Student exercise (solved): histogram version

```{r}
for (i in 3:ncol(df1)) {
  print(
    df1 |> 
      ggplot(aes(x = df1[, i], fill = CHAS)) +
      geom_histogram(alpha = 0.4, bins = 30) +
      labs(x = colnames(df1[i]))
  )
}
```

---

## Combining plots on a single page

```{r}
plot_list <- list()

for (i in 3:ncol(df1)) {
  plot_list[[i - 2]] <- df1 |> 
    ggplot(aes(x = df1[, i], color = CHAS)) +
    geom_density() +
    labs(x = colnames(df1[i]))
}

wrap_plots(plot_list, ncol = 2)
```

---

## Functional programming approach (solved)

```{r}
plot_list2 <- map(
  names(df1[3:ncol(df1)]),
  function(col_name) {
    ggplot(df1, aes(x = .data[[col_name]], color = CHAS)) +
      geom_density() +
      labs(x = col_name)
  }
)

wrap_plots(plot_list2, ncol = 2)
```

---

### Student exercise (solved): histogram + fill

```{r}
plot_list2_hist <- map(
  names(df1[3:ncol(df1)]),
  function(col_name) {
    ggplot(df1, aes(x = .data[[col_name]], fill = CHAS)) +
      geom_histogram(alpha = 0.4, bins = 30) +
      labs(x = col_name)
  }
)

wrap_plots(plot_list2_hist, ncol = 2)
```

---

## Scatter plots vs MEDV (solved)

```{r}
plot_list3 <- map(
  names(df1[3:ncol(df1)]),
  function(col_name) {
    ggplot(df1, aes(x = .data[[col_name]], y = MEDV)) +
      geom_point(aes(color = CHAS)) +
      labs(x = col_name)
  }
)

wrap_plots(plot_list3, ncol = 2)
```

---

### Student exercise (solved): add regression line

```{r}
plot_list4 <- map(
  names(df1[3:ncol(df1)]),
  function(col_name) {
    ggplot(df1, aes(x = .data[[col_name]], y = MEDV)) +
      geom_point(aes(color = CHAS)) +
      geom_smooth(method = "lm", se = FALSE) +
      labs(x = col_name)
  }
)

wrap_plots(plot_list4, ncol = 2)
```

---

## Box plots (solved)

```{r}
df1 |> 
  ggplot(aes(y = MEDV, x = CHAS)) +
  geom_boxplot()
```

---

## Homework – example solution

Below is **one acceptable solution** using a functional approach.

```{r}
boxplot_list <- map(
  names(df1[3:ncol(df1)]),
  function(col_name) {
    ggplot(df1, aes(x = CHAS, y = .data[[col_name]])) +
      geom_boxplot() +
      labs(y = col_name)
  }
)

wrap_plots(boxplot_list, ncol = 2)
```



---

## Key takeaways

* Programmatic plotting scales to high-dimensional data
* Functional programming integrates naturally with ggplot
* Visualization is a diagnostic step for modeling
* Clean EDA reduces downstream modeling errors


