---
title: "01-caret-basics"
format: html
editor: visual
---

## Caret basics: a unified modeling framework

### Learning objectives

After completing this module, you should be able to:

- Explain what the `caret` package abstracts and why it is useful
- Prepare outcome variables correctly for classification in caret
- Perform stratified train–test splits for classification problems
- Apply preprocessing correctly without data leakage
- Train classification models using cross-validation
- Evaluate and compare models using ROC-based metrics
- Tune model hyperparameters using grid search

---

## What is `caret`?

`caret` (Classification And REgression Training) is a **wrapper framework** that provides:

- a unified syntax for over 200 ML algorithms
- consistent handling of:
  - preprocessing
  - resampling
  - tuning
  - performance evaluation

Rather than learning dozens of model-specific APIs, we learn **one workflow**.

---

## Installing and loading data

The datasets used here come from the *Machine Learning for Business Analytics* textbook.

```{r}
#| eval: false
#| output: false
#| warning: false
library(devtools)

#install_github("gedeck/mlba/mlba", force = TRUE)
# or the forked version
#install_github("econjjacob/mlbadata", force = TRUE)
```

```{r}
#| output: false
#| warning: false
library(caret)
library(tidyverse)
library(lattice)
library(mlbadata)
```

---

## Data structure and outcome preparation

We use the **Universal Bank** dataset.

```{r}
bank.df <- mlbadata::UniversalBank
str(bank.df)
```

### Important caret rule

For **classification** in caret:

* the outcome **must be a factor**
* the **first level is the target class**

Here, `Personal.Loan` is numeric (`0` / `1`).
We convert it carefully.

```{r}
bank.df$Personal.Loan <- factor(
  bank.df$Personal.Loan,
  labels = c("No", "Yes")
) |> 
  relevel(ref = "Yes")
```

---

### Predictor cleanup

```{r}
bank.df <- bank.df |> 
  select(!c(ZIP.Code, ID)) |> 
  mutate(
    Education = factor(
      Education,
      levels = c(1, 2, 3),
      labels = c("UGrad", "Grad", "Adv/Prof")
    )
  )

str(bank.df)
```

---

### Student check

```{r}
#| eval: false
# Why must the outcome variable be a factor
# before using caret for classification?
```

---

## Data partitioning (stratified)

We reserve **20% for testing**, using stratified sampling.

```{r}
set.seed(1234)

trIndex <- createDataPartition(
  bank.df$Personal.Loan,
  p = 0.8,
  list = FALSE
)

train.df <- bank.df[trIndex, ]
test.df  <- bank.df[-trIndex, ]
```

Check class balance:

```{r}
print(paste0("Full data: ", mean(bank.df$Personal.Loan == "Yes")))
print(paste0("Training set: ", mean(train.df$Personal.Loan == "Yes")))
print(paste0("Test set: ", mean(test.df$Personal.Loan == "Yes")))
```

---

## Preprocessing without data leakage

**Rule:** preprocessing must be learned on the training set only.

### Explicit preprocessing approach

```{r}
train.vals <- preProcess(train.df, method = c("center", "scale"))

train.trnf <- predict(train.vals, train.df)
test.trnf  <- predict(train.vals, test.df)

glimpse(test.trnf)
```

Note:

* numeric variables were standardized
* factor variables were untouched

---

### Student check

```{r}
#| eval: false
# Why is it incorrect to estimate preprocessing
# parameters using the full dataset?
```

---

## Model training with `train()`

The `train()` function:

* performs resampling
* tunes hyperparameters
* selects the optimal model
* estimates expected performance

### Define resampling strategy

```{r}
set.seed(89)

fitCntrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  allowParallel = TRUE
)
```

---

## Basic logistic regression

```{r}
logistic1 <- train(
  Personal.Loan ~ Age + Income + CreditCard,
  data = train.trnf,
  method = "glm",
  family = "binomial",
  trControl = fitCntrl
)

logistic1
summary(logistic1)
```

---

## Using ROC as the optimization metric

Probability-based metrics require extra settings.

```{r}
fitCntrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 3,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

```{r}
logistic2 <- train(
  Personal.Loan ~ Age + Income + CreditCard,
  data = train.trnf,
  method = "glm",
  family = "binomial",
  trControl = fitCntrl,
  metric = "ROC"
)

logistic2
summary(logistic2)
```

---

## Test set evaluation

```{r}
test.trnf.pred <- predict(logistic2, newdata = test.trnf)

confusionMatrix(
  data = test.trnf.pred,
  reference = test.trnf$Personal.Loan
)
```

---

## In-place preprocessing with `train()`

Caret can perform preprocessing internally.

```{r}
logistic3 <- train(
  Personal.Loan ~ Age + Income + CreditCard,
  data = train.df,
  preProc = c("center", "scale"),
  method = "glm",
  family = "binomial",
  trControl = fitCntrl,
  metric = "ROC"
)

logistic3
```

```{r}
test.pred <- predict(logistic3, newdata = test.df)

confusionMatrix(
  data = test.pred,
  reference = test.df$Personal.Loan
)
```

Compare predictions:

```{r}
table(test.pred == test.trnf.pred)
```

---

## Hyperparameter tuning: penalized logistic regression

Elastic net combines:

* **ridge** (L2 penalty)
* **lasso** (L1 penalty)

```{r}
glmn_grid <- expand.grid(
  .alpha  = seq(0, 1, 0.1),
  .lambda = seq(0.01, 0.3, length = 20)
)

nrow(glmn_grid)
```

```{r}
glmnetMod <- train(
  Personal.Loan ~ .,
  data = train.df,
  preProc = c("center", "scale"),
  trControl = fitCntrl,
  method = "glmnet",
  tuneGrid = glmn_grid,
  metric = "ROC"
)

glmnetMod
```

---

## Model interpretation

```{r}
predictors(glmnetMod)
varImp(glmnetMod)
```

---

## Final evaluation

```{r}
glmnClass <- predict(glmnetMod, newdata = test.df)

confusionMatrix(
  data = glmnClass,
  reference = test.df$Personal.Loan
)
```

---

## Key takeaways

* caret enforces **good ML hygiene**
* outcome encoding matters
* preprocessing must avoid leakage
* cross-validation drives model selection
* ROC is preferred under class imbalance
* regularization controls model complexity



