---
title: "Regression Methods"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Regression

### Learning objectives

After completing this section, you should be able to:

* Fit a multiple linear regression model using training data
* Interpret regression output and goodness-of-fit statistics
* Generate out-of-sample predictions using a validation set
* Evaluate regression performance using RMSE and residual diagnostics
* Visually diagnose potential model issues using residual plots

---

## Load the required libraries

```{r}
library(tidyverse)
library(broom)
library(forecast)
library(caret)
library(mlbadata)
```

---

## Data preparation

We will use the **Toyota Corolla** dataset to predict used car prices.

To keep computation manageable and interpretation clear, we:

* use the **first 1,000 observations**
* select a subset of relevant variables
* treat fuel type as a categorical variable

```{r}
car.df <- mlbadata::ToyotaCorolla

# use first 1000 rows of data
car.df <- car.df[1:1000, ]

head(car.df)

# convert fuel type to factor
car.df$Fuel_Type <- as.factor(car.df$Fuel_Type)

head(car.df)
```

Select variables for regression:

```{r}
selected.var <- c(3, 4, 7, 8, 9, 10, 12, 13, 14, 17, 18)
car.df.sel <- car.df[, selected.var]
```

---

### Student exercise 1: predictors

```{r}
#| eval: false
# Which of the selected variables do you expect
# to have the strongest relationship with Price?
# Briefly explain why.
```

---

## Train–validation split

We partition the data into:

* **training set (60%)** → model estimation
* **validation set (40%)** → out-of-sample evaluation

```{r}
set.seed(1)

trainInd <- createDataPartition(
  car.df.sel$Price,
  p = 0.6,
  list = FALSE
)

train.df <- car.df.sel[trainInd, ]
valid.df <- car.df.sel[-trainInd, ]
```

---

### Student exercise 2: evaluation logic

```{r}
#| eval: false
# Why do we evaluate model performance
# on the validation set rather than
# the training set?
```

---

## Baseline multiple linear regression

We begin with a **full model**, including all selected predictors.

```{r}
lm.1 <- lm(Price ~ ., data = train.df)
```

Model output:

```{r}
tidy(lm.1)
```

Goodness-of-fit statistics:

```{r}
glance(lm.1)
glance(lm.1)$adj.r.squared
```

### Interpretation notes

* Coefficients describe *partial* effects holding other variables constant
* Adjusted (R^2) accounts for model complexity
* At this stage, we are **not** performing feature selection

---

### Student exercise 3: interpretation

```{r}
#| eval: false
# Identify one predictor with:
# (a) a statistically significant effect
# (b) an unexpected sign
# Provide a possible explanation.
```

---

## Out-of-sample prediction

We now evaluate the model on **unseen data**.

Rather than using `predict()` directly, we use `augment()` to:

* attach predictions
* retain residuals
* keep everything in a single dataframe

```{r}
head(augment(lm.1, newdata = valid.df))
```

Rename columns for clarity:

```{r}
augment(lm.1, newdata = valid.df) |> 
  rename(
    Pricehat = .fitted,
    resid = .resid
  ) -> valid.pred
```

---

## Model performance metrics

We compute standard regression accuracy measures.

```{r}
postResample(valid.pred$Pricehat, valid.df$Price)
```

### Interpretation

* RMSE measures average prediction error in original units
* These metrics reflect **generalization performance**

---

### Student exercise 4: metric interpretation

```{r}
#| eval: false
# Is RMSE easier to interpret than R-squared?
# Why might a business user prefer RMSE?
```

---

## Residual diagnostics

Residual plots help diagnose:

* nonlinearity
* heteroskedasticity
* outliers

### Residuals vs fitted values

```{r}
valid.pred |> 
  ggplot(aes(x = Pricehat, y = resid)) +
  geom_point()
```

### Distribution of residuals

```{r}
valid.pred |> 
  ggplot(aes(x = resid, y = ..density..)) +
  geom_histogram()
```

---

### Student exercise 5: diagnostics

```{r}
#| eval: false
# Based on the residual plots:
# (a) Do you see any patterns?
# (b) What might these patterns suggest
#     about model assumptions?
```

---

## Key takeaways

* Linear regression provides a strong baseline model
* Evaluation must be done on unseen data
* RMSE is a practical, interpretable metric
* Residual plots reveal model limitations
* This framework extends naturally to
  feature selection and regularization

```
