---
title: "10-01 Logistic Regression"
format: html
editor: visual
error: true
echo: true
---

## Learning objectives

After completing this module, students should be able to:

- Explain why linear regression is inappropriate for classification
- Interpret logistic regression coefficients as log-odds and odds ratios
- Fit logistic regression models using `glm()`
- Generate predicted probabilities from a logistic model
- Understand how probability cutoffs relate to classification decisions

---

## Ch 10: Logistic Regression

Classification refers to assigning a success or failure class (in the case of binary outcomes) to an observation. This class assignment is based on generating a probability of the outcome for that case and then classifying the record using a probability cutoff value.

Since we need to generate probabilities, we cannot use traditional ordinary least squares regression: there is nothing in OLS that restricts predictions to lie between 0 and 1. Logistic regression addresses this issue by modeling probabilities through a nonlinear response function.

This file introduces logistic regression using the base R `glm()` syntax.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(caret)
library(yardstick)
library(broom)
library(mlbadata)
````

---

## Why linear regression fails for classification

Let us first model the acceptance of a personal loan as a function of income.

Let

* ( p ) = probability of accepting a personal loan

A linear probability model is given by:

[
p = \beta_0 + \beta_1 \cdot income
]

The **logistic response function** is:

[
p = \frac{1}{1 + \exp^{-(\beta_0 + \beta_1 \cdot income)}}
]

We will estimate both models and visualize the difference.

---

## Toy example: linear vs logistic fit

### Data setup

```{r}
toy.df <- mlbadata::UniversalBank[, c(4, 10)]

toy.df$Class <- factor(
  toy.df$Personal.Loan,
  levels = c(0, 1),
  labels = c("No", "Yes")
)

glimpse(toy.df)
```

### Linear regression (incorrect model)

```{r}
reg.1 <- lm(Personal.Loan ~ Income, data = toy.df)
tidy(reg.1)
```

### Logistic regression (correct model)

```{r}
logit.1 <- glm(Class ~ Income, family = "binomial", data = toy.df)
tidy(logit.1)
```

---

## Comparing fitted curves

```{r}
lin.mod <- function(x) {
  reg.1$coefficients[1] + reg.1$coefficients[2] * x
}

logit.mod <- function(x) {
  1 / (1 + exp(-(logit.1$coefficients[1] + logit.1$coefficients[2] * x)))
}

toy.df %>% 
  ggplot(aes(x = Income, y = Personal.Loan)) +
  geom_jitter(width = 0.01, height = 0.02, alpha = 0.1) +
  geom_function(fun = lin.mod, color = "blue") +
  geom_function(fun = logit.mod, color = "red") +
  labs(
    x = "Income (in $000s)",
    y = "Personal Loan"
  )
```

---

### 🧠 Student Exercise 1 — Conceptual check

```{r}
#| eval: false

# Why is the blue (linear) curve problematic as a probability model?
# List two concrete issues visible in the plot.
```

---

## Logistic regression: general form

In general, the logistic response function is:

[
p = \frac{1}{1 + \exp^{-(\beta_0 + \beta_1 x_1 + \dots + \beta_K x_K)}}
]

Define the **odds of success** as:

[
Odds(Y = 1) = \frac{p}{1 - p}
]

Taking logs:

[
\log(odds) = \beta_0 + \beta_1 x_1 + \dots + \beta_K x_K
]

This log-odds formulation is what `glm()` estimates when `family = "binomial"`.

---

## Full logistic regression model

### Data preparation

```{r}
bank.df <- mlbadata::UniversalBank
bank.df <- bank.df[, -c(1, 5)]  # Drop ID and ZIP code

bank.df$Education <- factor(
  bank.df$Education,
  levels = c(1, 2, 3),
  labels = c("Undergrad", "Grad", "Adv/Prof")
)

bank.df$Personal.Loan <- factor(
  bank.df$Personal.Loan,
  levels = c(0, 1),
  labels = c("No", "Yes")
)

glimpse(bank.df)
```

### Train–validation split

```{r}
set.seed(1111)

train.index <- createDataPartition(
  bank.df$Personal.Loan,
  p = 0.8,
  list = FALSE
)

train.df <- bank.df[train.index, ]
valid.df <- bank.df[-train.index, ]
```

---

## Estimating the logistic regression

```{r}
logit.reg <- glm(
  Personal.Loan ~ .,
  data = train.df,
  family = "binomial"
)

summary(logit.reg)
tidy(logit.reg)
```

---

### 🧠 Student Exercise 2 — Interpretation

```{r}
#| eval: false

# Pick one continuous predictor.
# How do you interpret its coefficient in terms of:
# (a) log-odds
# (b) odds
```

---

## Odds ratios

To aid interpretation, we exponentiate the coefficients.

```{r}
result.1 <- tidy(logit.reg, exponentiate = TRUE)

result.1 %>% 
  mutate(
    estimate = round(estimate, 2),
    p.value = round(p.value, 4)
  )
```

A value greater than 1 indicates that an increase in the predictor increases the odds of being in class "Yes".

---

### 🧠 Student Exercise 3 — Odds ratios

```{r}
#| eval: false

# Identify one predictor with an odds ratio greater than 1
# and one with an odds ratio less than 1.
# Explain what each means in plain language.
```

---

## Predicted probabilities

Predictions from a logistic regression are **probabilities**, not classes.

```{r}
trainpred.df <- augment(
  logit.reg,
  data = train.df,
  type.predict = "response"
)

head(trainpred.df)
```

Alternatively:

```{r}
logit.reg.trpred <- predict(
  logit.reg,
  type = "response"
)

head(logit.reg.trpred)
```

---

### 🧠 Student Exercise 4 — Classification logic

```{r}
#| eval: false

# Suppose we use a cutoff of 0.5.
# How would predicted probabilities be converted into class labels?
# What happens if we lower the cutoff to 0.25?
```

---

## Wrap-up

Logistic regression models probabilities through a nonlinear transformation that ensures valid predictions between 0 and 1. Coefficients are interpreted through log-odds and odds ratios, and classification decisions depend critically on the choice of probability cutoff.

In the next module, we will evaluate logistic models using confusion matrices, ROC curves, and lift charts.

