---
title: "04-01-summ-cor"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Cleaning

A first step to any analysis is examining the descriptive statistics from the dataframe and so some preliminary pre-processing.

Load the required libraries

```{r}
library(tidyverse)
library(caret)
library(corrplot)
library(psych)
library(hablar)
library(skimr)
```


Load the dataset and convert two variable to factors. We will learn how to use the `apply()` family of functions which apply functions in r to various elements of a list.

```{r}
housing.df <- read.csv("BostonHousing.csv", stringsAsFactors = F) 

# convert CHAS and CAT..MEDV to factors

cols <- c("CHAS", "CAT..MEDV")
housing.df[cols] <- lapply(housing.df[cols], factor, levels = c("1", "0"), 
                           labels = c("Yes", "No"))

```


There are two factor variables in the above dataset: CHAS and CAT..MEDV. We first create a vector with these two columns and then apply (list apply) the factor function to the list of these two function. The syntax of `lapply()` funtion is:

```
lapply(X, FUN, …)
```
The `...` means that various arguments from the chosen `FUNC` can be passed here. Thus, we first made 1 as the first level and then addedd the labels Yes and No for 1 and 0 respectively.

```{r}
head(housing.df)
```

Looking at descriptive stats:

```{r}
summary(housing.df)
# a better format
skim(housing.df)

# customize the output
my_skim <- skim_with(numeric = sfl(p25 = NULL, p75 = NULL)) 
#sfl: skim function list
my_skim(housing.df)

```


We can also create pivot tables. Suppose we want to see average median price by RAD and Charles River.

```{r}
housing.df |> 
  group_by( RAD, CHAS) |> 
  summarise( Average.Value = mean(MEDV),
             count = n())
```


Another critical aspect of pre processing is creating factor variables and interactions. Here, we have two dummy variables, CHAS and CAT..MEDV. Let us create a dummy variable for CHAS and the interaction of CHAS and ROOMS

```{r}
df1 <- dummyVars(~ CHAS + CHAS:RM, data = housing.df,
                 fullRank = T)

housing.df1 <- predict(df1, newdata = housing.df)
```

We can now create a new dataframe where we combine the new dataset with housing.df

```{r}

housing.df2 <- cbind(housing.df, housing.df1)
housing.df2 <- housing.df2 |> 
  select(-CHAS)
```


## Dimension reduction: Corrlelation analysis

Having a high correlation between predictor variables can result in multicollinearity in regression based models. Before any modelling is undertaken. If two variables are highly correlated (say, above a certain threshold like 0.8), the one which is highly correlated with other variables should be removed.

`caret` has a built in function, `findCorrelation()` which follows this algorithm in removing such highly correlated variables. 

We will first apply some data cleaning methods. We will separate out our dependent variables (MEDV and CAT..MEDV) into a separate dataset, remove categorical variables from the the xvariable dataset and then compute correlations.

```{r}
yvars <- housing.df |> 
  select(MEDV, CAT..MEDV)

xvars <- housing.df |> 
  select(-MEDV, -CAT..MEDV)

xvarsnum <- xvars |> 
  select_if(is.numeric) #CHAS will get ommited
# if there were multiple categorical variables, we could have pulled them as !na.numeric or is.factor

cor1 <- cor(xvarsnum)
cor1

corrplot(cor1)
corrplot(cor1, method = "number")
# organized by clusters of high correlation
corrplot(cor1, order = "hclust" , method = "number")

# This method is cool but gets hard to read with several variables.
library(GGally)
ggpairs(xvarsnum[, c(1, 3, 4, 6)])
```

We do see above that there are some variable with rather high pair wise correlation. Let's identify and remove them.

```{r}
high.corr <- findCorrelation(cor1, cutoff = 0.8)
high.corr

t(names(xvarsnum[high.corr]))

#var is TAX. Next we remove it
xvarfin <- xvarsnum[, -high.corr]
```

Combine this dataset with CHAS and MEDV

```{r}
df1 <- cbind(MEDV = yvars$MEDV, CHAS = xvars$CHAS, xvarfin)
```

Some prelim plots with selected variables:

```{r}
selected.vars <- c("MEDV", "CHAS" ,"RM", "ZN", "PTRATIO", "AGE", "CRIM")

df2 <- df1[, selected.vars]
for (i in 3:ncol(df2)) {
  print(df2 |> 
    ggplot(aes(x = df2[,i], color = CHAS)) +
    geom_density() +
    labs(x = colnames(df2[i])))
    }

for (i in 3:ncol(df2)) {
  print(df2 |> 
    ggplot(aes(y = MEDV, x = df2[,i])) +
    geom_point() +
    geom_smooth() +
    labs(x = colnames(df2[i])))
     }
#box plots work with categorical x variables
  df2 |> 
    ggplot(aes(y = MEDV, x = CHAS)) +
    geom_boxplot() 
```

